---
title: "Data 607 Final Project"
author: "Coco Donovan"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction:

I chose to have the focus of my analysis be police incident reports from the City 
of San Francisco Police Department. My intention was to include analysis of crime 
clearance statistics, and unfortunately that is not possible and I will get into 
that dynamic a little during my presentation. Ultimately, I chose to conduct an 
analysis over the years of crime and reports of crime.

### Limitations: 

There is no readily available data for before 2003, so I cannot make quantitative 
statements from my own analysis of crime trends from before 2003 (I would love to
based on reporting I have seen; however that would not be my own work).

There are two different data sets; one from 2003-2018 and another from 2018-current.
From this change in data sources has come a difference in values for the same 
variables (this makes certain forms of analysis difficult to produce).

### Necessary Packages

```{r}
library(httr)
library(jsonlite)
library(ggplot2)
library(dplyr)
library(readr)
library(tidyverse)
```

### Police Incident Data: 2003 - 2017

```{r}
old_police_files  <- list.files(pattern = '^Police_Data_....\\.csv')

old_police_tables <- lapply(old_police_files, read.csv, header = TRUE)

old_police_data <- do.call(rbind , old_police_tables)
```

### Police Incident Data: 2018 - Current

```{r}
present_police_files  <- list.files(pattern = 'Present_Police_Data_....\\.csv')

present_police_tables <- lapply(present_police_files, read.csv, header = TRUE)

present_police_data <- do.call(rbind , present_police_tables)
```

### SF Yearly Population

```{r}
sf_pop <- read.csv('SF_POP.csv')

sf_pop$Population <- 1000 * sf_pop$Population

sf_pop <- sf_pop %>%
  separate(Year, c('year', 'month', 'date'), '-') %>%
  select(year, Population)

sf_pop$year <- as.integer(sf_pop$year)

sf_pop <- sf_pop %>%
  filter(year >= 2003)
```

### Joining New and Old Incident Counts, Reports, and the ratio of the two:

```{r}
old_police_data <- old_police_data %>%
  separate(Date, c("month", "day","year"), "/")

old_police_data$year <- as.integer(old_police_data$year)

old_counts_and_incidents <- old_police_data %>%
  group_by(year) %>%
  summarize(count_reports = as.double(n()), count_incidents = as.double(length(unique(IncidntNum))))

new_counts_and_incidents <- present_police_data %>%
  rename('year' = 'Incident.Year') %>%
  group_by(year) %>%
  summarize(count_reports = as.double(n()), count_incidents = as.double(length(unique(Incident.Number))))

counts_and_incidents <- rbind(new_counts_and_incidents, old_counts_and_incidents)

counts_and_incidents <- counts_and_incidents %>%
  group_by(year) %>%
  mutate(ratio = count_reports/count_incidents) %>%
  arrange(desc(year))

knitr::kable(counts_and_incidents)
```

### Visualizations: Incidents Over Time

```{r}
# I did not include 2023 just yet, because the year is only a about 1/3 of the 
# way over and would not contribute to reliable comparisions for raw counts

counts_and_incidents_no_2023 <- subset(counts_and_incidents, year != 2023)

ggplot(counts_and_incidents_no_2023, aes(x= year, y= count_incidents)) +
  geom_point() +
  geom_line() +
  labs(x = 'Year', y = 'Incidents per Year', title = 'Incidents per Year Since 2003')
```

### Visualization: Ratio of reports vs acutal incidents

```{r}
ggplot(counts_and_incidents, aes(x= year, y= ratio)) +
  geom_point() +
  geom_line() +
  labs(x = 'Year', y = 'Ratio of Reports Filed v.s. Actual Incidents', title = 'The Ratio of Reported Crimes to Actual Incidents Since 2003')
```

### Adding population to counts and incidents

```{r}
years_and_counts <- full_join(sf_pop, counts_and_incidents, by = 'year')
```

### Crime per 100K

```{r}
years_and_counts <- years_and_counts %>%
  mutate(reports_per_100k = 100000 * count_reports/Population, incidents_per_100k = 100000 * count_incidents/Population) %>%
  arrange(desc(year))

knitr::kable(years_and_counts)
```

### Visualization: Crime Incidents per 100k

```{r}
ggplot(years_and_counts, aes(x= year, y= incidents_per_100k)) +
  geom_point() +
  geom_line() +
  labs(x = 'Year', y = 'Incidents per Year per 100k Residents', title = 'Incidents per Year per 100k Residents Since 2003')
```

### Arrest Rate Rate

This is not possible for over time comparison and that has to do with how SF records 
"Resolutions" to incidents. There is no easily accessible (I could not find any after
scouring SFPD's website and SF City and County Documentation) documentation on what 
each resolution type meant for the older data, and then the classifications were 
changed with no explanation. Due to this lack of transparency, it is difficult to 
track crime clearance rates over time. It does seem that this is data that does 
exist internally and it is not open sourced.

```{r}
new_outcomes <- present_police_data %>%
  group_by(Resolution) %>%
  summarize(counts = n())

unique(new_outcomes$Resolution)

old_outcomes <- old_police_data %>%
  group_by(Resolution) %>%
  summarize(count = n())

unique(tolower(old_outcomes$Resolution))
```

### Incident Code Crosswalk:

I am going to have to do some manual work to make a full cross walk because if
just use the DataSF crosswalk that leaves me with about 6800 obersvations without 
a crime category. Not to mention, the missing crime categories occur more often 
in the earlier years which would slightly skew results in the direction of the 
dangerous SF narrative.

```{r}
# This incident codes crosswalk comes from DataSF and using it leads to missing 
# crime categories. Essentially, this crosswalk is not sufficient and needs to be 
# improved and the process will unfortunately be largely manual.

#incident_codes <- jsonlite::fromJSON('https://data.sfgov.org/resource/ci9u-8awy.json?$limit=1012')

#incident_codes$inc_code <- as.integer(incident_codes$inc_code)

#incident_codes <- incident_codes %>%
#  rename(Incident.Code = inc_code, Category = category, Subcategory = subcategory)

incident_codes <- present_police_data %>%
  select(Incident.Code, Incident.Category) %>%
  unique()

# I could continue this process, but this gets the number of observations missing
# a category down to a little less the 800, which is significantly better than the 
# 6793 observations missing a category when you use DataSF's proprietary crosswalk
# and truthfully this is a bit of tedious process.

row1 = c(Incident.Code = 11010, Incident.Category = 'Stolen Property')
row2 = c(Incident.Code = 28090, Incident.Category = 'Vandalism')
row3 = c(Incident.Code = 30090, Incident.Category = 'Other Miscellaneous')
row4 = c(Incident.Code = 0, Incident.Category = 'Other Miscellaneous')
row5 = c(Incident.Code = 6311, Incident.Category = 'Larceny Theft')
row6 = c(Incident.Code = 16644, Incident.Category = 'Drug Violation')

incident_codes <- rbind(incident_codes, row1, row2, row3, row4, row5, row6)

incident_codes$Incident.Code <- as.integer(incident_codes$Incident.Code)

# Combined Drug Offense and Drug Violation into one category

incident_codes$Incident.Category[incident_codes$Incident.Category == 'Drug Offense'] <- 'Drug Violation'

# lumped the blank category in with miscellaneous

incident_codes$Incident.Category[incident_codes$Incident.Category == ''] <- 'Other Miscellaneous'

# The manual process of finding the missing categories:

old_incident_codes <- old_police_data %>%
  select(Incident.Code, Category) %>%
  filter()

unique(old_incident_codes$Category)

is_it_in_old <- old_incident_codes %>%
  filter(Incident.Code == 'insert integer value Incident.Code')

categories <- unique(incident_codes$Incident.Category)
```

### Analysis of Crime Type:

```{r}
type_old <- old_police_data %>%
  select(year, IncidntNum, Incident.Code) %>%
  rename(Incident.Number = IncidntNum, Incident.Year = year)

type_new <- present_police_data %>%
  select(Incident.Year, Incident.Number, Incident.Code)

total_type <- bind_rows(type_old, type_new)

incidents_with_types <- left_join(total_type, incident_codes)

# -------------------------------
# Checking the missing values:

dist_missing_types <- incident_codes %>%
  filter(is.na(Incident.Category)) %>%
  #group_by(Incident.Year) %>%
  summarise(n())

missing_types <- incident_codes %>%
  filter(is.na(Incident.Category)) %>%
  group_by(Incident.Code) %>%
  summarize(n())
# ------------------------------

counts_by_category <- incidents_with_types %>%
  group_by(Incident.Year, Incident.Category) %>%
  summarise(count_reports = as.double(n()), count_incidents = as.double(length(unique(Incident.Number)))) %>%
  mutate(ratio = count_reports/count_incidents) %>%
  arrange(Incident.Year, desc(count_reports))

counts_by_category <- counts_by_category %>%
  rename(year = Incident.Year)

year_pop_counts_categories <- full_join(sf_pop, counts_by_category, by = 'year')

counts_by_category <- year_pop_counts_categories %>%
  group_by(year, Incident.Category) %>%
  mutate(incidents_per_100k = 100000 * count_incidents/Population) %>%
  arrange(year, desc(count_reports))
```

### Visualization: Drug Violations

```{r}
counts_by_category_drug_violations <- counts_by_category %>%
  filter(Incident.Category == 'Drug Violation')

# Overall Incidents over Time

ggplot(counts_by_category_drug_violations, aes(x= year, y= count_incidents)) +
  geom_point() +
  geom_line() +
  labs(x = 'Year', y = 'Drug Violations per Year', title = 'Drug Violations per Year Since 2003')

# Incidents Per 100k

ggplot(counts_by_category_drug_violations, aes(x= year, y= incidents_per_100k)) +
  geom_point() +
  geom_line() +
  labs(x = 'Year', y = 'Drug Violations per Year per 100k Residents', title = 'Drug Violations per Year per 100k Residents Since 2003')

# Ratio of Reports vs Actual Incidents

ggplot(counts_by_category_drug_violations, aes(x= year, y= ratio)) +
  geom_point() +
  geom_line() +
  labs(x = 'Year', y = 'Ratio of Reports Filed v.s. Actual Incidents', title = 'The Ratio of Reported Crimes to Actual Incidents Since 2003')
```

### Visualization: Larceny Theft

```{r}
counts_by_category_larceny <- counts_by_category %>%
  filter(Incident.Category == 'Larceny Theft')

# Overall Incidents over Time

ggplot(counts_by_category_larceny, aes(x= year, y= count_incidents)) +
  geom_point() +
  geom_line() +
  labs(x = 'Year', y = 'Larceny Theft Incidents per Year', title = 'Larceny Theft Incidents per Year Since 2003')

# Incidents Per 100k

ggplot(counts_by_category_larceny, aes(x= year, y= incidents_per_100k)) +
  geom_point() +
  geom_line() +
  labs(x = 'Year', y = 'Larceny Theft Incidents per Year per 100k Residents', title = 'Larceny Theft Incidents per Year per 100k Residents Since 2003')

# Ratio of Reports vs Actual Incidents

ggplot(counts_by_category_larceny, aes(x= year, y= ratio)) +
  geom_point() +
  geom_line() +
  labs(x = 'Year', y = 'Ratio of Reports Filed v.s. Actual Incidents', title = 'The Ratio of Reported Crimes to Actual Incidents Since 2003')
```

### Visualization: Rape

```{r}
counts_by_category_rape <- counts_by_category %>%
  filter(Incident.Category == 'Rape')

# Overall Incidents over Time

ggplot(counts_by_category_rape, aes(x= year, y= count_incidents)) +
  geom_point() +
  geom_line() +
  labs(x = 'Year', y = 'Rapes per Year', title = 'Rapes per Year Since 2003')

# Incidents Per 100k

ggplot(counts_by_category_rape, aes(x= year, y= incidents_per_100k)) +
  geom_point() +
  geom_line() +
  labs(x = 'Year', y = 'Rapes per Year per 100k Residents', title = 'Rapes per Year per 100k Residents Since 2003')

# Ratio of Reports vs Actual Incidents

ggplot(counts_by_category_rape, aes(x= year, y= ratio)) +
  geom_point() +
  geom_line() +
  labs(x = 'Year', y = 'Ratio of Reports Filed v.s. Actual Incidents', title = 'The Ratio of Reported Crimes to Actual Incidents Since 2003')
```

### Homicide

It is extremely hard to track homicide with DataSF and SFPD data. The data on
homicide only goes back to 2018, not to mention and incident does not necessarily
translate to a singular homicide and rather could translate to multiple homicides
at one time. So, given the homicides are a rarer form of crime and the fact that
one incident can actually mean multiple deaths sort of mystifies this already small
set of results. Also, manslaughter is classified under the general homicide umbrella,
likely necessitating a sub-category analysis on an already small category. I know
accurate Homicide data exists under SFPD's ownership, but it is not readily available
to the public.

```{r}
counts_by_category_homicide <- counts_by_category %>%
  filter(Incident.Category == 'Homicide')

knitr::kable(counts_by_category_homicide)
```


### SF City Budget

A continuation of this analysis could use this file that comes from DataSF detailing
the cities budget over the years, with a specific focus on funding of the Police.

```{r}
budget <- read.csv('Budget.csv')
```

### Conclusion:

The trends that I observed were as follows:

The incidents of crime have gone up yearly, over the timeframe, with the exception 
of the pandemic. However, when we look at crime per 100K residents, crime has 
gone down generally, with a large increase during the pandemic. 

What is especially of note is that as crime per 100k residents, the reporting of
crime increased. It is difficult to comment on exactly why that is because of 
lack of clearance rates and the inconsistency with the "Resolution" variable in 
both police incident report data sets.
